---
title: "Tutorial_solutions_pmb"
authors: "Jonah Thomas, Gilberte Martine-Edith, Petra M. Baumann"
format: 
  html:
    toc: true
    theme: spacelab
    axe:
      output: console
editor: visual
---

# Solutions

### Exercise 1: Abbott LibreView (AL001)

```{r, eval=FALSE}
data_AL001 <- read.csv("data/AL001_CGM_data_2days.csv", skip = 2, header = TRUE)
head(data_AL001[1:5])
```

### Exercise 2: Dexcom data from Glooko (DG002)

```{r, eval=FALSE}
data_DG002 <- read.csv("data/DG002_CGM_data_2days.csv", skip = 1, header = TRUE)
head(data_DG002)
```

### Exercise 3: Dexcom Clarity (DC003)

```{r, eval=FALSE}
data_DC003 <- read.csv("data/DC003_CGM_data_2days.csv")
head(data_DC003)

data_DC003 <- read.csv("data/DC003_CGM_data_2days.csv", sep = ";")
head(data_DC003)
ncol(data_DC003)
```

### Exercise 4: Medtronic CareLink (MC004)

```{r, eval=FALSE}
data_MC004 <- read.csv("data/MC004_CGM_data_2days.csv", skip=6, sep=";")
head(data_MC004)
```

### Exercise 6: Reading GluCINDa output into R

```{r, eval=FALSE}
glucinda_data <- read.csv("data2_parsed/parsed_just_cgm.csv", sep=";")
head(glucinda_data)
colnames(glucinda_data)
str(glucinda_data)
```

### Exercise 7: applying the conversion function

```{r, eval=FALSE}
glucinda_data <- glucinda_data %>% 
  add_converted_glucose_columns()
  

summary(glucinda_data$value_mmoll)
```

### Exercise 8: verify the datetime column

```{r, eval=FALSE}
class(glucinda_data$datetime)   # expected: "POSIXct" "POSIXt"
head(glucinda_data$datetime)
```

### Exercise 9: Fixing duplicate values

```{r, eval=FALSE}
iglu_data <- iglu_data %>% dplyr::group_by(id, time) %>% summarise(gl = mean(gl))

duplicates <- iglu_data %>% dplyr::filter( duplicated(paste(id, time)) | duplicated(paste(id, time), fromLast = TRUE) )

nrow(duplicates)
```

### Exercise 10: Calculate mean and range of glucose values per participant

```{r, eval=FALSE}
iglu_data_14days %>% 
  group_by(id) %>% 
  summarise(mean_glucose = mean(gl, na.rm = TRUE),
            min_glucose = min(gl, na.rm = TRUE), 
            max_glucose = max(gl, na.rm = TRUE))
```

### Exercise 11: CGM metrics

```{r,eval=FALSE}
iglu::active_percent(iglu_data_14days)
iglu::mean_glu(iglu_data_14days)
iglu::cv_glu(iglu_data_14days)
iglu::ea1c(iglu_data_14days)
iglu::above_percent(iglu_data_14days)
iglu::in_range_percent(iglu_data_14days)
iglu::below_percent(iglu_data_14days)
```

### Exercise 12: Visualizing CGM data

```{r,eval=FALSE}

iglu::plot_glu(iglu_data_14days)

iglu::agp(iglu_data_14days %>% dplyr::filter(id == "AL001"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "DG002"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "DC003"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "MC004"))
```
