---
title: "CGM Data Analysis Tutorial"
authors: "Jonah Thomas, Gilberte Martine-Edith, Petra M. Baumann"
format: 
  html:
    toc: true
    theme: spacelab
    axe:
      output: console
editor: visual
---

# Step 1: Downloading data

This process will be discussed as part of the presentation - there is no need to use R or any coding languages to download data. Instead you should follow the process which will be specific to each CGM device/brand.

For the tutorial, we will use two sets of example data from <https://github.com/leicester-cdag/CGMTraining>. Download the `Tutorial.qmd`, `data` and `data2` folders and save them in a new folder. `data` and `data2` contain the same files, but `data/` only covers two days of CGM and includes an additional example file. We will start with `data/`.

# Step 2: Reading CGM data into R

## 1. Installing packages

Before doing any data cleaning or analysis in R, you need to install and load the appropriate packages that will allow you to run codes smoothly. For this tutorial, we will need the `dplyr`, `lubridate`, and `iglu` packages.

```{r}
install.packages("dplyr")
install.packages("lubridate")
install.packages("iglu")
```

```{r}
library(dplyr)
library(lubridate)
library(iglu)
```

## 2. Reading a single file

Many different file formats can be read into R using a variety of functions. CSV and Excel files are likely to be the file types most commonly read. CSV files can be read into R using the `read.csv()` function whilst Excel files require the `readxl` package and `read_xlsx()` function. The `haven` package is extremely useful for reading in less common files formats such as SPSS or Stata files.

Most functions which read data into R take a file path as the first argument; this specifies which file you want to read into R and where it is saved. Below is an example of how to read a csv file in R.

```{r, eval=TRUE}
firstimport <- read.csv("data/First_import_exampledata.csv")
head(firstimport)
```

`head()` shows the first six rows. You should see column names on the first row and data immediately below.

## 3. When the defaults are not enough

There is no standard for CGM exports which is why table formats differ significantly between CGM manufacturers. Many include metadata rows above the actual column names, which means `read.csv()` will misread the file unless you tell it to skip those rows.

Therefore, your need two more arguments in `read.csv()`:

| Argument | What it does |
|----|----|
| `skip = n` | Skips the first *n* lines before reading, default is 0 |
| `header = TRUE` | Treats the next line after skipping as column names (default) |

### Exercise 1: Abbott LibreView (AL001)

Read in `AL001_CGM_data_2days.csv`. Open the raw file first in a text editor or spreadsheet application and count how many rows appear before the actual column names, then set `skip` and `header` accordingly. Name the object `data_AL001` and then look at the first few rows.

```{r}
# Write your code for Exercise 1 below



```

::: callout-caution
The dataframe looks okay for now, but note that the file currently has no column telling you the participant ID whose data you just imported. We will discuss this later.
:::

### Exercise 2: Dexcom data from Glooko (DG002)

Read in `DG002_CGM_data_2days.csv`. **Hint:** This Dexcom export from Glooko has one metadata row above the column names. Name the object `data_DG002`.

```{r}
# Write your code for Exercise 2 below


```

## 4. When skipping is not enough

### Exercise 3: Dexcom Clarity (DC003)

Try reading `DC003_CGM_data_2days.csv` without any extra arguments. Name the object `data_DC003`

```{r, eval=TRUE}
# Write your code for Exercise 3 below


```

No error, but check the number of columns and look at the output carefully. Here are some more useful functions to help you explore a dataframe.

```{r, eval=TRUE}
nrow(data_DC003) # Number of rows
ncol(data_DC003) # Number of columns
str(data_DC003)  # Structure of an object 
```

The data frame contains a single column with all values crammed together as one string. If you open the raw file in a text editor, you will see that the values are not separated by commas but by a different character. This is a common failure mode: the file reads without error, the result looks like a data frame, but the data is not actually parsed. Therefore, always check column count and column names before proceeding.

You will need another argument in the `read.csv()` function to solve this problem:

| Argument    | What it does                                            |
|-------------|---------------------------------------------------------|
| `sep = ","` | Character used to separate columns; default is a comma. |

Use the same object name (`data_DC003`) to overwrite the previous import. Explore the first rows and the number of columns.

```{r, eval=TRUE}
# Write your code for Exercise 3 (2nd try) below


```

Even after fixing the separator, something still looks off. Let's click table icon next to `data_DC003` in the **Environment** pane in RStudio to open the full data viewer and scroll through the rows.

The first several rows after the column names contain metadata and device settings - noticeable, even though the data are in German, because there is no date/time information. This means that even a correctly separated file can require additional cleaning before the data are usable. You could filter out rows with not date/time information, but the column that holds the timestamp has a different name in every language and different manufacturer's exports, so the filtering logic you write for DC003 will not transfer to the next file. You are accumulating bespoke fixes, not building a pipeline. Let's look at one last example before discussing what to do instead.

### Exercise 4: Medtronic CareLink (MC004)

Now try `MC004_CGM_data_2days.csv`. Looking at the raw data, we see that the separator is semicolon and we need to skip 6 rows.

```{r, eval=TRUE}
# Your code for Exercise 4

```

No error, but click the viewer and scroll down. Medtronic CareLink exports contain multiple sections (pump settings, aggregated statistics, and sensor readings), each preceded by a divider and a new header row. Let's confirm this:

```{r, eval=TRUE}
str(data_MC004)
```

Because of the intermediate headers, there are string values in all columns, which is why they all got recognized as character, not as numeric/integer like in the other files. This means you cannot do calculations on the glucose column right now.

To fix this, you would have to filter rows containing `"-------"` or `"Index"` in the `Index` column and then change the column type manually. Sound exhausting?

## 5. The problem at scale

We have just seen four files from three manufacturers and four download platforms, each requiring a different approach. In a multi-centre study you might receive exports from multiple systems simultaneously, across different software versions, language settings, and regional date formats. Writing and maintaining bespoke parsing code for each combination is fragile, time-consuming, and produces undocumented transformations that are difficult to reproduce or audit.

Luckily, there is a better solution:

# GluCINDa

**GluCINDa** (*Glucose: Coherent Import. Neat Data.*) is an open-source Python tool designed to solve exactly this problem. It acts as a **manufacturer-agnostic ingestion layer**: it reads raw CGM exports from multiple manufacturers in a single run, applies transparent and documented parsing rules, and produces a single harmonised output file with a stable, analysis-ready schema.

## Installing GluCINDa

We will install GluCINDa now from PyPI using `pip`. You can do that in RStudio's terminal (middle tab in the bottom R Studio pane) by typing:

``` bash
pip install glucinda
```

OR

``` bash
py -m pip install --user glucinda
```

If you are using a Mac device, try the following to install gluCINDa:

```         
python3 -m pip install --user glucinda
```

## Launching GluCINDa

Once installed, launch the graphical user interface by typing in RStudio's terminal:

``` bash
glucinda
```

OR

``` bash
py -m glucinda
```

If you are using a Mac device, try the following to launch gluCINDa:

``` bash
python3 -m glucinda
```

Take a few minutes to explore the interface before proceeding.

## Using GluCINDa

### Exercise 5: Running GluCINDa on your data

::: callout-important
GluCINDa creates a column containing the participant ID which is read from the filename. The only requirement for GluCINDa is, therefore, that the CGM export **file names start with** **subject ID (3-20 characters)** **followed by an underscore** (`_)`.

The files for this tutorial are named correctly already.
:::

In the GluCINDa GUI:

1.  Set the **input folder** to the `data2/` directory containing the regular-length CGM export files.
2.  Use the automatically suggested **default** **output folder**.
3.  Click **Run**.

GluCINDa will process all files it recognises and write several output files to the output folder, including HTML summary reports.

::: callout-tip
This process may take a few minutes. Explore the description of outputs in the README at <https://github.com/GluCINDa> in the meantime.
:::

GluCINDa will display a clickable link to the output directory once the process is finished. The file we need for the rest of this tutorial is **`parsed_just_cgm.csv`**: one row per CGM measurement, across all input files and manufacturers, in a single consistent schema, pre-filtered to only contain the regular CGM stream.

::: callout-note
A pre-processed parsed_just_cgm.csv is available in `glucinda_output/` in case you encounter installation issues. Continue from the next section using that file.
:::

### Exercise 6: Reading GluCINDa output into R

You can now import the GluCINDa data with `read.csv()`like above. Note that GluCINDa uses the semicolon as separator and make sure to read \`data2_parsed/parsed_just_cgm.csv\`. Call the object \`glucinda_data\` and inspect the resulting dataframe

```{r, eval=TRUE}
# Write your code for Exercise 6 below


```

The key columns for our purposes are:

| Column | Content |
|----|----|
| `sid` | Subject/participant identifier, derived from filename |
| `date` | Date of measurement (`YYYY/mm/dd`) |
| `time` | Time of measurement (`HH:MM:SS`) |
| `value_num` | Numeric glucose value |
| `unit` | Unit of measurement as it appears in the source column header (e.g. `mg/dl`, `mmol/l`); `undefined` if absent |
| `manufacturer_assumed` | CGM device manufacturer inferred from the input file structure |
| `source_file` | Original filename — full traceability to raw data |

### Sanity checks

Let's have a quick look at the reports produced by GluCINDa before proceeding.

We can also explore individual columns of the imported dataframe:

```{r, eval=TRUE}

table(glucinda_data$manufacturer_assumed)
table(glucinda_data$source_file)

```

------------------------------------------------------------------------

# Step 3: Formatting columns

The iglu package is commonly used to analyse CGM data within the diabetes research community. The package contains many functions which allow us to summarise and visualise CGM data in a variety of ways. We will be using `iglu` from Step 5 onwards but before doing so we need to prepare the harmonized data frame, working through five steps:

1.  Check glucose values (unit, range, missingness)
2.  Create `value_mgdl` — only after confirming units
3.  Build unit conversion functions and create `value_mmoll`
4.  Parse and combine the date and time columns
5.  Select and rename to the `iglu` format (`id`, `time`, `gl`)

## 1. Checking glucose values

We inspect the data before making any transformations. A problem found before transformation is straightforward to address; the same problem discovered afterwards may be ambiguous in origin.

### Unit column

```{r, eval=FALSE}
table(glucinda_data$unit)
```

For this dataset you should see only `"mg/dL"`. In a multi-centre study you might see both units; you would need to harmonise them before proceeding — the function we write in Section 3 handles exactly this.

### Glucose values

```{r, eval=TRUE}
summary(glucinda_data$value_num)
```

CGM sensors operate within a physiological detection range, typically 40–400 mg/dL (approximately 2.2–22.2 mmol/L). Values outside this window indicate sensor errors or artefacts and warrant investigation.

Let's check the missing values. These are probably all rows where GluCINDa extracted a string value from a glucose column. We now also introduce the magrittr pipe operator \`%\>%\`:

```{r, eval=TRUE}
glucinda_data %>% 
  dplyr::filter(is.na(value_num)) %>% 
  dplyr::group_by(source_file, sid, value_str) %>% 
  dplyr::summarise(n=n()) %>% mutate(n_total = sum(n))
```

We will keep the string values for now because deleting them would skew the calculation of sensor activity.

## 2. Creating `value_mgdl`

Only once you have confirmed that all numeric values in this dataset are in mg/dL, is it safe to create a new variable using the [`mutate` function](https://dplyr.tidyverse.org/reference/mutate.html) from the dplyr package:

```{r, eval=FALSE}
glucinda_data <- glucinda_data  %>% 
  dplyr::mutate(value_mgdl = value_num)
```

But, if your dataset contained a mix of units, you would need to convert based on the unit. For that purpose, we create a custom function. Note that this will overwrite the variable `value_mgdl`we just created.

## 3. Unit conversion: Introducing functions

Good analysis code avoids embedding raw numeric constants directly in expressions without explanation. We name our conversion constants first:

```{r, eval=TRUE}
mgtommol_multiplier <- 0.0555
mmoltomg_multiplier <- 18.018018
```

We can now write a **function** built on these constants. A function in R is a named piece of code that takes inputs (*arguments*) and returns a result. You define it once and call it wherever you need it.

The function below uses `ifelse` to decide, for each row, whether the value is already in the target unit or needs to be converted. The condition is checked using `grepl("mg", unit)`, which searches for a substring (first argument) in a column (second argument) and returns `TRUE` or `FALSE` per row. In this case, if `TRUE`, the value is in mg/dL and is passed through unchanged; if `FALSE`, the multiplier is applied. The second call searches for `"mol"` and applies the multiplier in case it is not found. Note that we are looking for a substring which is more robust than exact matching because it does not depend on the precise string. This way, both "mg/dl" and "mg/dL" would be identified.

```{r, eval=TRUE}
add_converted_glucose_columns <- function(mydata) {
  mydata  %>% 
    dplyr::mutate(
      value_mgdl  = ifelse(grepl("mg", unit),
                           value_num,
                           value_num * mmoltomg_multiplier),
      value_mmoll = ifelse(grepl("mol", unit),
                           value_num,
                           value_num * mgtommol_multiplier)
    )
}
```

### Exercise 7: applying the conversion function

Applying a function means passing an object (such as a dataset) into it to produce an output. See an example of defining then applying a function below:

```{r, eval=FALSE}

# Define function
classify <- function(data) {
  data %>% 
    dplyr::mutate(newcolumn = 
                    ifelse(oldcolumn >= 7, 
                           "High", 
                           "Normal")

}

# Option 1: Apply the function using a pipe
dataset <- dataset %>%
  classify()

# Option 2: Apply the function not using a pipe
dataset <- classify(dataset)

```

Now, with the help of the examples above, apply `add_converted_glucose_columns()` to the `glucinda_data` dataset.

```{r}
# Write your code for Exercise 7 below



```

You now have both `value_mgdl` and `value_mmoll`, derived from named constants and a documented function which is fully traceable and reproducible across any dataset you apply this code to.

## 4. Combining date and time character columns

GluCINDa stores date and time in two separate character columns. `iglu` requires a single date-time column. Run the code chunk below to firstly combine them into a single column using the `paste` function:

```{r, eval=TRUE}
glucinda_data <- glucinda_data  %>% 
  dplyr::mutate(
    datetime = paste(date, time)
  )
```

## 5. Changing date time character column format

The two columns have been combined into the `datetime` column, however they are still in a character format. For our analysis to run successfully, we need to change this to a date/time format. We can do that using base R `as.POSIXct` function as shown in the code below:

```{r,eval=FALSE}
data <- data |>
  dplyr::mutate(
    time = as.POSIXct(time, format = "%d/%m/%Y %H:%M", tz="UTC")
  )
```

Alternatively, you can also do this using the `lubridate` package, which provides intuitive and flexible [functions](https://lubridate.tidyverse.org/reference/ymd_hms.html) to format date-time columns.

### Exercise 8

Using this [help page](https://lubridate.tidyverse.org/reference/ymd_hms.html), define which function from the `lubridate` package would help you format your time column into the correct date time format and write the code below:

```{r}
# Insert your code for exercise 8 below 



```

## 5. Selecting columns for `iglu`

`iglu` expects a data frame with exactly three columns: `id`, `time`, and `gl` (glucose in mg/dL).

There are two main methods used to select columns in R - referred to as the base method or the dplyr (tidyverse) method. To select columns in base R we can select columns by their number as shown below.

```{r, eval=F}
data <- data[, 1] # select column 1 
data <- data[, 1:3] # select columns 1 through 3
data <- data[, c(1,2, 4:6)] # select columns 1 and 2, and 4 through 6
```

Using `dplyr` we can select columns using their names as shown below.

```{r, eval=F}
data <- data |>
  dplyr::select(column_1, column_2)
```

### Exercise 9

Create a new dataframe called `iglu_data` based on `glucinda_data` and try either of the approaches described above to select columns sid, datetime and value_mgdl:

```{r, eval=TRUE}
#Write your code for exercise 9 below

```

## Renaming columns for `iglu`

As with selecting columns, there are many options regarding how columns can be renamed. In base R, a simple method to rename columns is shown below.

```{r,eval=F}
colnames(data) <- c("new_name1", "new_name2")
```

In dplyr, we can use the `rename` function.

```{r,eval=F}
data <- data |>
  dplyr::rename(
    new_name = "old_name",
    new_name2 = "old_name2"
  )
```

### Exercise 10

Try to rename the columns of your `iglu_data` dataframe. Your new column names should be: id, time, gl.

```{r}
# Write your code for exercise 10 below




```

------------------------------------------------------------------------

# Step 4: Running checks on the data

CGM data is fundamentally a time series. Errors in timestamps — duplicates, readings outside the expected study window, or systematic gaps — will distort any derived metric, including TIR, TBR, and `active_percent`. These checks use the `datetime` column created in Steps 3.4 and 3.5. We run them on `iglu_data` rather than `glucinda_data` because the provenance columns `source_file` and `manufacturer_assumed` are still present there — if a problem is found, you need them to identify which input file or device it came from.

## 1. Duplicate timestamps

A duplicate timestamp means two readings exist for the same participant at the same moment. This should not occur and indicates a data integrity problem — for example, a file processed twice or a sensor-device overlap.

```{r, eval=TRUE}
duplicates <- iglu_data  %>% 
  dplyr::filter(
    duplicated(paste(id, time)) |
    duplicated(paste(id, time), fromLast = TRUE)
  )

nrow(duplicates)     # should be 0
duplicates     # inspect if not
```

For now, we will take the mean of the duplicate values as shown below. If we wanted to investigate further, the additional columns in `glucinda_data` may be helpful. Run the code below to calculate the mean glucose for the duplicated timestamps.

```{r, eval=TRUE}
iglu_data <- iglu_data %>%
  dplyr::group_by(id, time) %>%
  dplyr::summarise(gl = mean(gl)) %>%
  dplyr::ungroup()

```

## 2. Temporal range and activity

CGM sensor activity is calculated by comparing the expected number of measurements with the actual number. The expected number of measurements depends on the measurement interval:

-   every 5 minutes: 60/5=12 measurements per hour, or 12\*24=288 measurements per day per participant

-   every 15 minutes: 60/15=4 measurements per hour, or 4\*24=96 measurements per day per participant

### Exercise 11

Check the temporal range of each participant's data and sensor activity using the (`active_percent`)\[https://irinagain.github.io/iglu/reference/active_percent.html\] function from the `iglu` package:

```{r, eval=TRUE}
# Write your code for Exercise 11 below



```

What can you notice? Who has the highest number of CGM data collection days? Who has the lowest sensor activity percentage?

We only need 14 days of data for the tutorial, so let's select the last 14 full days (leaving out the last day per subject as these are likely truncated) for each participant to get data from the middle of the data. Run the code below:

```{r, eval=TRUE}
iglu_data_14days <- iglu_data %>% 
  dplyr::group_by(id) %>%
  dplyr::filter(
    as.Date(time) >= max(as.Date(time)) - 14,
    as.Date(time) <= max(as.Date(time)) - 1
    ) %>%
  dplyr::ungroup()

# Check that the runtimes are now only 14 days and activity is sufficient
iglu_data_14days %>% 
  iglu::active_percent()   
```

## 3. Value checks

We now want to check the range (min, max) of glucose values for each participant. How would you do this?

### Exercise 12: Calculate range of glucose values per participant

```{r, eval=TRUE}
# Your code for Exercise 12

```

# Step 5: Computing CGM metrics

You are now ready to output CGM-derived metrics using the `iglu` package. Explore the functionality available within the iglu package. You can do this by reading this [webpage](https://irinagain.github.io/iglu/reference/index.html) or by typing `iglu::` followed by a letter into the console in R.

## Exercise 13

We would like you to produce:

-   The mean glucose for each participant

-   The coefficient of variation for each participant

-   The estimated HbA1c for each participant

-   The time above, below, and in range for each participant

```{r, eval=TRUE}
# Your code for Exercise 13



```

# Step 6: Visualising CGM data using iglu

There are many ways we can plot CGM data. The simplest of these is a time series graph. This is a line graph with time on the X axis, and glucose value on the Y axis. It allows us to see all the data collected on a given participant, identify any periods of missing data, see the highest and lowest glucose values recorded, and do a basic quality check to identify any potential issues with our data.

The ambulatory glucose profile (AGP) is another commonly used graph to display an individual's average glucose level over a given period of time (often a week or 14 days). These plots allow us at a glance to assess an individual's glucose profile. `iglu` provides a function for plotting CGM data as an AGP which also includes a table with key CGM metrics and an in range percentage plot which tells us the percentage of time a participant spent below, above, and in range.

## Exercise 14

For this exercise we would like you to plot two graphs for a participant of your choice: a time series graph and an AGP. The `iglu` package has functions to create all these.

Once you have created this series of graphs for one participant, please create the same series of graphs for another participant. Using these graphs try to compare the glucose management between the two. What is similar, what is different, does anything stand out?

```{r, eval=TRUE}
# Write your code for exercise 14 below



```

# Final exercise

You should now have the skills to harmonize CGM data with GluCINDa, read them into R, manipulate into your desired format and use the `iglu` package to calculate your required metrics and visualisations. As a final exercise, create an R script (Go to `File -> New file... -> R Script`) with your full pipeline of codes.
