---
title: "Continuous Glucose Monitoring Data Analysis Group Training"
author: "Jonah Thomas and Gilberte Martine-Edith"
format: 
  html:
    toc: true
    theme: spacelab
    axe:
      output: console
editor: visual
---

# Introduction

This course will teach you how to process, analyse, and visualise CGM data using R. We will break this process down into 6 steps: downloading data, reading CGM data into R, formatting columns (data manipulation), running checks, outputting CGM metrics, and visualising CGM metrics.

## Getting help with R

There are many ways to get help with coding in R. If you want additional help within RStudio, you can search for a function by running `?dplyr::select` in the R console with any function. This is helpful to understand how a function is intended to work and what arguments it may take. For larger (more complex) questions, ChatGPT, Microsoft CoPilot, and other AI tools can be extremely helpful (but can be prone to errors). There are also a number of forums online (including StackOverflow) where your question may previously have been answered.

## Exercises and solutions

Throughout this document, there will be several coding exercises where you will be asked to write some code to solve a problem. A solution to each exercise can be found at the end of this document. We recommend trying to write the code yourself first, then trying to get help from the internet or Rstudio, then asking your colleagues sat around you, then asking Gil or myself, before consulting the solution. However, they are always available if you need a nudge in the right direction.

# Step 1: Downloading data

This process will be discussed as part of the presentation - there is no need to use R or any coding languages to download data. Instead you should follow the process which will be specific to each CGM device/brand.

# Step 2: Reading data into R

## 1. Installing packages

Before doing any data cleaning or analysis in R, you need to install and load the appropriate packages that will allow you to run codes smoothly. For this tutorial, we will need the `dplyr`, `lubridate`, and `iglu` packages. Run the codes below to install and load these packages.

```{r}
##Installing packages 
install.packages("dplyr")
install.packages("lubridate")
install.packages("iglu")

##Loading packages 
library(dplyr)
library(lubridate)
library(iglu)
```

## 2. Reading a single file

Many different file formats can be read into R using a variety of functions. CSV and Excel files are likely to be the file types most commonly read. CSV files can be read into R using the `read.csv` function whilst Excel files require the `readxl` package and `read_xlsx` file. The `haven` package is extremely useful for reading in less common files formats such as SPSS or Stata files.

Most functions which read data into R take a file path as the first argument; this specifies which file you want to read into R and where it is saved. Below is an example of how to read a csv file in R.

```{r, eval=FALSE}
data <- read.csv("C:/Users/gme6/OneDrive - University of Leicester/CGM Training/test.csv")
```

### Exercise 1

Now it is your turn to read a file into R. We would like you to read in the csv file titled "Patient A" in your folder into RStudio using the code above as an example. Once you have read the file into R, view it either using the `View()` command or clicking on it in the environment.

```{r}
# Your code to complete exercise 1
```

### An additional challenge

You may notice when looking at your data that something is not quite right. The column names we want are currently stored in our first row of data. This issue can be resolved using specific arguments you can add to the `read.csv` function. Specifically, see what the `skip` and `header` arguments do when they are changed.

```{r}
# Your additional code can go here
```

## 3. Reading multiple files at once

Very often when working with CGM data it is necessary to read in multiple files (often one per participant). In large studies, this would require repeating our above code hundreds of times. To reduce this repetition, we can use a programming concept known as a "for loop". A "for loop" allows us to execute the same lines of code multiple times while iteratively changing aspects of the code. In our example, we can change the file path to read in a different file each time.

Before reading in multiple files we need to get a list of the file names into R. We can do this by setting a folder path (where our CGM data is stored), and then using the `list.files` function to get the names of the files in that folder.

```{r,eval=F}
folder_path <- "C:/Users/gme6/OneDrive - University of Leicester/CGM Training"
files <- list.files(folder_path)
```

Once we have our list of files, we can read the data in using a for loop and save it as a list of dataframes.

```{r,eval=F}
data_list <- list() 

for(i in 1:length(files)) {
  data_list[[i]] <- read.csv(paste0(folder_path, "/", files[i]), skip = 1, header = TRUE)
}

names(data_list) <- files

data <- dplyr::bind_rows(data_list, .id = "id") |>
  dplyr::mutate(id = gsub(".csv", "", id))
```

In the above code, we create an empty list to save our data into. We then create a for loop which will iterate through from number 1, to the number of files stored in the folder. We then use the `paste0` function to create a file path by pasting together the folder path with the file name. This read in data is there stored in the list (data_list).

Once we have our list of dataframes, we then use the `names` function to name each dataframe. In this case we have named them the same as the original file names, but this can be changed to suit your needs. We use the `dplyr` function `bind_rows` which creates one large, long dataframe where all our data is now stored. We use the `.id` argument to add an ID column to our data (so we know which participant the data came from) which in this scenario will be the file names. Lastly, the function `mutate` slightly modifies the column id so that ".csv" is removed from participant's ID.

The above code is relatively complex to understand and if it does not all make sense straight away that is to be expected. Fortunately, this code can be copy and pasted into any project and by adjusting the folder path can be run without any further adjustments.

## Exercise 2

Copy the code from above which allows you to read in multiple files. Change the folder path to the folder where you have saved your CGM data. Once this is changed, the code should run and read in the files contained within the folder.

```{r}
# Your exercise 2 code can go here
```

# Step 3: Formatting columns

Once our data is read into R, we often need to do some basic data manipulation to get the data in the format we require. One step in this process is selecting the appropriate columns, then changing the column names.

## 1. Selecting columns

There are two main methods used to select columns in R - referred to as the base method or the dplyr (tidyverse) method. To select columns in base R we can select columns by their number as shown below.

```{r, eval=F}
data <- data[, 1] # select column 1 
data <- data[, 1:3] # select columns 1 through 3
data <- data[, c(1,2, 4:6)] # select columns 1 and 2, and 4 through 6
```

Using `dplyr` we can select columns using their names as shown below.

```{r, eval=F}
data <- data |>
  dplyr::select(column_1, column_2)
```

## Exercise 3

Try either of the approaches above to select the ID, date/time, and glucose columns in your dataset.

```{r}
# Insert your code to exercise 3 here 
```

## 2. Changing column names

As with selecting columns, there are many options regarding how columns can be renamed. In base R, a simple method to rename columns is shown below.

```{r,eval=F}
colnames(data) <- c("new_name1", "new_name2")
```

In dplyr, we can use the `rename` function.

```{r,eval=F}
data <- data |>
  dplyr::rename(
    new_name = "old_name",
    new_name2 = "old_name2"
  )
```

## Exercise 4

Try to rename the columns of your dataframe. They should be named: id, time, gl.

```{r}
# Your code for exercise 4 should go here
```

## 3. Changing date time columns to different formats

Columns which express the date and time are notoriously difficult to manage when coding! However, we do not need to do too much with our CGM data. Often when CGM data are read into R, they are read as a character column, but for our analysis to run successfully we need to change this to a date/time format. We can do that using base R `as.POSIXct` function as shown in the code below:

```{r,eval=FALSE}
data <- data |>
  dplyr::mutate(
    time = as.POSIXct(time, format = "%d/%m/%Y %H:%M", tz="UTC")
  )
```

You can also do this using the `lubridate` package, which provides intuitive and flexible [functions](https://lubridate.tidyverse.org/reference/ymd_hms.html) to format date-time columns.

## Exercise 5

Using this [help page](https://lubridate.tidyverse.org/reference/ymd_hms.html), define which function from the `lubridate` package would help you format your time column into the correct date time format and write the code below:

```{r}
# Insert your code to exercise 5 here 
```

## 4. Changing the format of the glucose column

Frequently in Europe, we look at CGM data measured in mmol/L. However, in the US there is a tendency to use mg/dL. The iglu package, which we will use later, only accepts glucose data in mg/dL. We therefore need to convert out glucose column from mmol/L to mg/dL. To do this, we need to multiply the column by 18.

## Exercise 6

See if you can write code to convert your glucose column from mmol/L, to mg/dL.

```{r}
# Your exercise 6 code goes here
```

# Step 4: Running checks on our data

Now we have our data read in and formatted correctly, we can run a few basic checks. This may include calculating the mean glucose, range of glucose values, and missingness within the CGM timestamps.

## Exercise 7

Conduct some basic data checks on your CGM data. Consider how you would calculate the:

-   Mean glucose value

-   Range of glucose values

-   How many CGM measurements would you expect compared with how many you have (each participant should have around 14 days of day)

```{r}
# Your code to address exercise 7 should go here 
```

# Step 5: Computing CGM metrics using the iglu package

The iglu package is commonly used to analyse CGM data within the research community. The package contains many functions which allow us to summarise and visualise CGM data in a variety of ways.

## Exercise 8

Explore the functionality available within the iglu package. You can do this by reading this [webpage](https://irinagain.github.io/iglu/reference/index.html) or by typing `iglu::` followed by a letter into the console in R.

We would like you to produce:

-   \% CGM active

-   The mean glucose for each participant

-   The coefficient of variation for each participant

-   The estimated HbA1c for each participant

-   The time above, below, and in range for each participant

```{r}
# Your code to solve exercise 8 should go here
```

# Step 6: Visualising CGM data using iglu

There are many ways we can plot CGM data. The simplest of these is a time series graph. This is a line graph with time on the X axis, and glucose value on the Y axis. It allows us to see all the data collected on a given participant, identify any periods of missing data, see the highest and lowest glucose values recorded, and do a basic quality check to identify any potential issues with our data.

The ambulatory glucose profile (AGP) is another commonly used graph to display an individual's average glucose level over a given period of time (often a week or 14 days). These plots allow us at a glance to assess an individual's glucose profile. `iglu` provides a function for plotting CGM data as an AGP which also includes a table with key CGM metrics and an in range percentage plot which tells us the percentage of time a participant spent below, above, and in range.

## Exercise 9

For this exercise we would like you to plot two graphs: a time series graph and an AGP for one participant. The iglu package has functions to create all these. Once you have created this series of graphs for one participant, please create the same series of graphs for another participant. Using these graphs try to compare the glucose management between the two. What is similar, what is different, does anything stand out?

```{r}
# Your solution for exercise 9 goes here
```

# Conclusion

You should now have the skills to read CGM data into R, manipulate the data into your desired format, use the iglu package to calculate your required metrics and visualisations. The final exercise below asks you to combine all the code you have used above to create a single pipeline which takes you from your csv file, all the way through to CGM metrics and an AGP visualisation.

```{r}
# Your final CGM data pipeline goes here
```

# Solutions

## Exercise 1

```{r,eval=FALSE}
data <- read.csv("your_file_path")
View(data)
```

### An additional challenge solution

```{r,eval=FALSE}
data <- read.csv("your_file_path", skip = 1, header = TRUE)
View(data)
```

## Exercise 2

```{r,eval=F}
folder_path <- "your_file_path"
files <- list.files(folder_path)

data_list <- list() 

for(i in 1:length(files)) {
  data_list[[i]] <- read.csv(paste0(folder_path, "/", files[i]), skip = 1, header = TRUE)
}

names(data_list) <- files

data <- dplyr::bind_rows(data_list, .id = "id")
```

## Exercise 3

```{r}
data <- data[, c(1,4,6)]

data <- data |>
  dplyr::select(id, Device.Timestamp, Historic.Glucose.mmol.L)
```

## Exercise 4

```{r}
colnames(data) <- c("id", "time", "gl")

data <- data |>
  dplyr::rename(
    id = "id",
    time = "Device.Timestamp",
    gl = "Historic.Glucose.mmol.L"
  )
```

## Exercise 5

```{r}
data <- data |>
  dplyr::mutate(
    time = lubridate::dmy_hm(time)
  )
```

## Exercise 6

```{r,eval=FALSE}
# only run one of these options once!

# method using base R
data$gl <- as.numeric(data$gl * 18)

# method using tidyverse syntax
data <- data |>
  dplyr::mutate(
    gl = as.numeric(gl * 18)
  )
```

## Exercise 7

```{r,eval=F}

mean(data$gl, na.rm = TRUE)
range(data$gl, na.rm = TRUE)

(14 * 5) * 96 # how many data points you would expect (14 days x 5 participants x 96 measurements (1440 / 15))
(14 * 5) * 96 - nrow(data)
```

## Exercise 8

```{r,eval=F}
iglu::active_percent(data)
iglu::mean_glu(data)
iglu::cv_glu(data)
iglu::ea1c(data)
iglu::above_percent(data)
iglu::in_range_percent(data)
iglu::below_percent(data)
```

## Exercise 9

```{r solution-8,eval=F}

iglu::plot_glu(data |> dplyr::filter(id == "Patient A"))
iglu::plot_glu(data |> dplyr::filter(id == "Patient B"))

iglu::agp(data |> dplyr::filter(id == "Patient A"))
iglu::agp(data |> dplyr::filter(id == "Patient B"))
```

## Final pipeline

```{r}
folder_path <- "your_file_path"
files <- list.files(folder_path)

data_list <- list() 

for(i in 1:length(files)) {
  data_list[[i]] <- read.csv(paste0(folder_path, "/", files[i]), skip = 1, header = TRUE)
}

names(data_list) <- files

data <- dplyr::bind_rows(data_list, .id = "id") |>
  dplyr::mutate(id = gsub(".csv", "", id))

data <- data[, c(1,4,6)]

colnames(data) <- c("id", "time", "gl")

data$gl <- as.numeric(data$gl * 18)

summary_metrics <- iglu::summary_glu(data)

iglu::glu(data |> dplyr::filter(id == "Patient A"))

iglu::agp(data |> dplyr::filter(id == "Patient A"))
```
