---
title: "Tutorial solutions"
authors: "Jonah Thomas, Gilberte Martine-Edith, Petra M. Baumann"
format: 
  html:
    toc: true
    theme: spacelab
    axe:
      output: console
editor: visual
---

# Solutions

### Exercise 1: Abbott LibreView (AL001)

```{r, eval=FALSE}
data_AL001 <- read.csv("data/AL001_CGM_data_2days.csv", skip = 2, header = TRUE)
head(data_AL001[1:5])
```

### Exercise 2: Dexcom data from Glooko (DG002)

```{r, eval=FALSE}
data_DG002 <- read.csv("data/DG002_CGM_data_2days.csv", skip = 1, header = TRUE)
head(data_DG002)
```

### Exercise 3: Dexcom Clarity (DC003)

```{r, eval=FALSE}
data_DC003 <- read.csv("data/DC003_CGM_data_2days.csv")
head(data_DC003)

data_DC003 <- read.csv("data/DC003_CGM_data_2days.csv", sep = ";")
head(data_DC003)
ncol(data_DC003)
```

### Exercise 4: Medtronic CareLink (MC004)

```{r, eval=FALSE}
data_MC004 <- read.csv("data/MC004_CGM_data_2days.csv", skip=6, sep=";")
head(data_MC004)
```

### Exercise 6: Reading GluCINDa output into R

```{r, eval=FALSE}
glucinda_data <- read.csv("data2_parsed/parsed_just_cgm.csv", sep=";")
head(glucinda_data)
colnames(glucinda_data)
str(glucinda_data)
```

### Exercise 7: applying the conversion function

```{r, eval=FALSE}
#Option 1
glucinda_data <- glucinda_data %>% 
  add_converted_glucose_columns()

#Option 2
glucinda_data <- add_converted_glucose_columns(glucinda_data)


summary(glucinda_data$value_mmoll)
```

### Exercise 8: Changing date time character column format

```{r, eval=FALSE}
glucinda_data <- glucinda_data %>%
  dplyr::mutate(datetime = lubridate::ymd_hms(datetime))
```

### Exercise 9: Selecting relevant columns

```{r, eval=FALSE}
#With base R

iglu_data <- glucinda_data[, c(1,19,21)]

#With dplyr

iglu_data <- glucinda_data  %>% 
  dplyr::select(sid, datetime, value_mgdl)

```

### Exercise 10: Renaming columns

```{r,eval=FALSE}
#With base R

colnames(iglu_data) <- c("id", "time", "gl")

#With dplyr

iglu_data <- iglu_data  %>% 
  dplyr::rename(id = sid, 
                time = datetime, 
                gl = value_mgdl)
```

### Exercise 11: Checking temporal range and sensor activity

```{r,eval=FALSE}
iglu::active_percent(iglu_data)
```

### Exercise 12: Calculate range of glucose values per participant

```{r,eval=FALSE}
#Example for Participant AL001
min(iglu_data %>% 
      dplyr::filter(id = "AL001"))
max(iglu_data %>% 
      dplyr::filter(id = "AL001"))
```

### Exercise 13: CGM metrics

```{r,eval=FALSE}
iglu::active_percent(iglu_data_14days)
iglu::mean_glu(iglu_data_14days)
iglu::cv_glu(iglu_data_14days)
iglu::ea1c(iglu_data_14days)
iglu::above_percent(iglu_data_14days)
iglu::in_range_percent(iglu_data_14days)
iglu::below_percent(iglu_data_14days)
```

### Exercise 14: Visualising CGM data

```{r,eval=FALSE}

iglu::plot_glu(iglu_data_14days)

iglu::agp(iglu_data_14days %>% dplyr::filter(id == "AL001"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "DG002"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "DC003"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "MC004"))
```


### Final exercise

```{r,eval=FALSE}

glucinda_data <- read.csv("data2_parsed/parsed_just_cgm.csv", sep=";") %>% 
  dplyr::mutate(value_mgdl = value_num)

mgtommol_multiplier <- 0.0555
mmoltomg_multiplier <- 18.018018

add_converted_glucose_columns <- function(mydata) {
  mydata  %>% 
    dplyr::mutate(
      value_mgdl  = ifelse(grepl("mg", unit),
                           value_num,
                           value_num * mmoltomg_multiplier),
      value_mmoll = ifelse(grepl("mol", unit),
                           value_num,
                           value_num * mgtommol_multiplier)
    )
}

glucinda_data <- glucinda_data %>% 
  add_converted_glucose_columns() %>% 
  dplyr::mutate(datetime = paste(date, time)) %>%
  dplyr::mutate(datetime = lubridate::ymd_hms(paste(date, time)))

iglu_data <- glucinda_data  %>%
  dplyr::select(sid, datetine, value_mgdl) %>%
  dplyr::rename(id=sid, time = datetime, gl = value_mgdl) %>%
  dplyr::group_by(id, time) %>%
  dplyr::summarise(gl = mean(gl)) %>%
  dplyr::ungroup()

iglu_data_14days <- iglu_data %>% 
  dplyr::group_by(id) %>%
  dplyr::filter(as.Date(time) >= min(as.Date(time)) + 20,
                as.Date(time) <= min(as.Date(time)) + 33) %>%
  dplyr::ungroup()

iglu::active_percent(iglu_data_14days)
iglu::mean_glu(iglu_data_14days)
iglu::cv_glu(iglu_data_14days)
iglu::ea1c(iglu_data_14days)
iglu::above_percent(iglu_data_14days)
iglu::in_range_percent(iglu_data_14days)
iglu::below_percent(iglu_data_14days)

iglu::plot_glu(iglu_data_14days)

iglu::agp(iglu_data_14days %>% dplyr::filter(id == "AL001"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "DG002"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "DC003"))
iglu::agp(iglu_data_14days %>% dplyr::filter(id == "MC004"))

```